They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-16 paper/paper.Rmd
some re-writing

Chance [-@chance2020] re-examines the constant probability notion in Major League Baseball's World Series (1923--2018), the National Basketball Association's Finals (1951--2018), and the National Hockey League's Stanley Cup (1939--2018). Chance finds strong evidence against the null hypothesis of $p = 0.5$ in the MLB and NHL championship series when applying Mosteller's first and second methods.[^8] Chance's work is closely related to mine, and, in fact, provides a guide for the first part of my investigation.

[^8]: Chance goes on to outline a conditional probability framework (likelihood of winning a game given the series state) which can exactly explain the distribution of the number of games played.

Momentum, one of most discussed topics in sports analytics, goes hand-in-hand with a discussion of the nature of series outcomes.[^9] Two opposing fallacies are observed in the context of momentum: the "gambler's fallacy" (negative recency) and "hot hand fallacy" (positive recency). Per Ayton et al. [-@ayton2004], negative recency is "the belief that, for random events, runs of a particular outcome ... will be balanced by a tendency for the opposite outcome", while positive recency is the expectation of observing future results that match recent results.

[^9]: We often use use "streaks" and momentum interchangeably, but as [@steeger2021] note, momentum implies dependence between events, whereas streaking does not.

Studying both player streaks and team streaks in basketball, in both observational and controlled settings. Gilovich et al. [-@gilovich1985] do not find evidence for the hot hand phenomenon. Recently, Miller et al. [-@miller2018] refuted the conclusions of Gilovich et al. , finding mathematical evidence that seems to support negative recency. Specifically, they find that a "bias exists in a common measure of the conditional dependence of present outcomes on streaks of past outcomes in sequential data" (streak selection bias) that imply that, under i.i.d. conditions, "the proportion of successes among the outcomes that immediately follow a streak of consecutive successes is expected to be strictly less than the underlying (conditional) probability of success". I agree with Miller's findings, accounting for the streak selection bias in my study of momentum.

Despite the plethora of existing research on games played in a series and momentum in sports, these topics have yet to be investigated heavily in esports. Work has been done to examine in-round win probability in other FPS titles such as Counter-Strike [@xenopoulos2022] and Valorant [@derover2021], both of which are round-based like CoD SnD. However, research on round-level trends is sparse, perhaps because games like Counter-Strike and Valorant both have economic aspects that can create clear advantages on side in a given round, given how prior rounds played out.[^10]

[^10]: Additionally, both Counter-Strike and Valorant have overtime rules and blocked offensive/defensive roles (i.e. playing either offense or defense for many consecutive rounds).

# Methodology, results, and discussion

First, I investigate the constant probability assumption and the distribution of rounds played in a series. Afterwards, I investigate momentum, building on my learnings from the constant probability assumption analysis.

@@

The general formula for the probability $P_E(r)$ that a best-of-$s$ series lasts $r$ rounds given constant probability $p$ of one team[^11] winning each round is

[^11]: If $p > 0.5$, then I might say that this team is the better team (known omnisciently).

@@

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
\caption{Alternate estimates of the constant probability \(p\) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive \(p = 0.5\).}\label{tbl:mosteller-methods-results} \\
\toprule()
Method & \(p(r)\) & \(\chi^2\) (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & \(p(r)\) & \(\chi^2\) (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum \(\chi^2\) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $P_E(r)$ when re-applying Equation \ref{eq:series-length} for the maximum likelihood estimate $p_2 = 0.575$.[^12] I observe that $P_E(r)$ is notably larger for $p_{2}$ when $r \in [6, 7]$, more closely matching $P_O(r)$. $p_{2}$ is also closer to $P_O(r)$ for $r \in [9, 10]$, although not for $r \in [8, 11]$.

[^12]: The method of moments and minimum $\chi^2$ estimates for $p$ are omitted simply because the results would be nearly identical to those for the maximum likelihood estimate of $p$ (since they are all $\approx 0.575$.

@@

While one might think that $E[\hat{p}^{kr,\text{win}} - \hat{p}^{kr,\text{loss}}] = 0$, Miller et al. [-@miller2018] suggest that, given a deterministic i.i.d. Bernoulli probability (Equation \ref{pwkr}, this is not true due to streak-selection bias.[^13][^14]

[^13]: Note that these probabilities are separate from those in Section \ref{sec:analysis-1}.

@@

I don't attempt to measure this expectation like Miller et al. [-@miller2018], who perform their analysis longitudinally per study participant; nonetheless, I apply their formulation for applying the "downward bias" on post-streak success to come up with estimates $\hat{P}_{kr,\text{win}}^{\text{adj.}}$, specifying $r$ as the number of trials.

Given that people typically perceive streaks as beginning after the third success (or failure) at minimum [@carlson2007], I focus on round win percentages in rounds immediately following streaks of three wins.[^15] Table \ref{tbl:cod-round-win-prop-after-k-wins} shows $\hat{P}_{kr,\text{win}}$ for $k = 3$ winning streaks, for both teams that empirically won and lost the series. (Some combinations of $r$ and $\text{team loses series}$ are irrelevant, such as the case when $r \in [7, 8, 9]$, so these combinations are excluded.)

[^15]: Three happens to also be a reasonable number for series that lasts at maximum 11 rounds.

@@

>>>>

Chance [-@chance2020] re-examines the constant probability notion in Major League Baseball's World Series (1923--2018), the National Basketball Association's Finals (1951--2018), and the National Hockey League's Stanley Cup (1939--2018). Chance finds strong evidence against the null hypothesis of $p = 0.5$ in the MLB and NHL championship series when applying Mosteller's first and second methods.[\^8] [\^8]: Chance goes on to outline a conditional probability framework (likelihood of winning a game given the series state) which can exactly explain the distribution of the number of games played.

Momentum, one of most discussed topics in sports analytics, goes hand-in-hand with a discussion of the nature of series outcomes.[^8] Two opposing fallacies are observed in the context of momentum: the "gambler's fallacy" (negative recency) and "hot hand fallacy" (positive recency). Per Ayton et al. [-@ayton2004], negative recency is "the belief that, for random events, runs of a particular outcome ... will be balanced by a tendency for the opposite outcome", while positive recency is the expectation of observing future results that match recent results.

[^8]: We often use use "streaks" and momentum interchangeably, but as [@steeger2021] note, momentum implies dependence between events, whereas streaking does not.

Studying both player streaks and team streaks in basketball, in both observational and controlled settings. Gilovich et al. [-@gilovich1985] do not find evidence for the hot hand phenomenon. However, Miller and Sanjurjo [-@miller2018] refuted the conclusions of Gilovich et al., proving a framework for quantifying streak selection bias, which effectively works in the manner posited by the gambler's fallacy. Specifically, Miller and Sanjurjo say that a "bias exists in a common measure of the conditional dependence of present outcomes on streaks of past outcomes in sequential data" implying that, under i.i.d. conditions, "the proportion of successes among the outcomes that immediately follow a streak of consecutive successes is expected to be strictly less than the underlying (conditional) probability of success". When applying streak selection bias to Giloveih et al.'s data, Miller and Sanjurjo come to the opposite conclusions as Giloveih et al.

Despite the plethora of existing research on games played in a series and momentum in sports, these topics have yet to be investigated heavily in esports. Work has been done to examine in-round win probability in other FPS titles such as Counter-Strike [@xenopoulos2022] and Valorant [@derover2021], both of which are round-based like CoD SnD. However, research on round-level trends is sparse, perhaps because games like Counter-Strike and Valorant both have economic aspects that can create clear advantages on side in a given round, given how prior rounds played out.[^9]

[^9]: Additionally, both Counter-Strike and Valorant have overtime rules and blocked offensive/defensive roles (i.e. playing either offense or defense for many consecutive rounds).

# Methodology and results

First, I investigate the constant probability assumption and the distribution of rounds played in a series. Chance's [-@chance2020] work is closely related to mine, and, in fact, provides a guide for this investigation.

Afterwards, I investigate post-streak win rates, building on my learnings from the constant probability assumption analysis. I agree with Miller and Sanjurjo's [-@miller2018] framework for streak selection bias, accounting for it in when analyzing momentum.

@@

The general formula for the probability $P_E(r)$ that a best-of-$s$ series lasts $r$ rounds given constant probability $p$ of one team[^10] winning each round is

[^10]: If $p > 0.5$, then I might say that this team is the better team (known omnisciently).

@@

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
\caption{Alternate estimates of the constant probability \(p\) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive \(p = 0.5\).}\label{tbl:mosteller-methods-results} \\
\toprule()
Method & \(p(r)\) & \(\chi^2\) (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & \(p(r)\) & \(\chi^2\) (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum \(\chi^2\) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $P_E(r)$ when re-applying Equation \ref{eq:series-length} for the maximum likelihood estimate $p_2 = 0.575$.[^11] I observe that $P_E(r)$ is notably larger for $p_{2}$ when $r \in [6, 7]$, more closely matching $P_O(r)$. $p_{2}$ is also closer to $P_O(r)$ for $r \in [9, 10]$, although not for $r \in [8, 11]$.

[^11]: The method of moments and minimum $\chi^2$ estimates for $p$ are omitted simply because the results would be nearly identical to those for the maximum likelihood estimate of $p$ (since they are all $\approx 0.575$.

@@

While one might think that $E[\hat{p}^{kr,\text{win}} - \hat{p}^{kr,\text{loss}}] = 0$, Miller and Sanjurjo [-@miller2018] suggest that, given a deterministic i.i.d. Bernoulli probability (Equation \ref{pwkr}, this is not true due to streak-selection bias.[^12][^13]

[^12]: Note that these probabilities are separate from those in Section \ref{sec:analysis-1}.

[^13]: Three happens to also be a reasonable number for series that lasts at maximum 11 rounds.

I apply their formulation for "downward bias" to post-streak success to come up with estimates $\hat{P}_{kr,\text{win}}^{\text{adj.}}$, specifying $r$ as the number of trials.

Given that people typically perceive streaks as beginning after the third success (or failure) at minimum [@carlson2007], I focus on round win percentages in rounds immediately following streaks of three wins.[^14] Table \ref{tbl:cod-round-win-prop-after-k-wins} shows $\hat{P}_{kr,\text{win}}$ for $k = 3$ winning streaks, for both teams that empirically won and lost the series. (Some combinations of $r$ and $\text{team loses series}$ are irrelevant, such as the case when $r \in [7, 8, 9]$, so these combinations are excluded.)

@@

@@

# Appendix {.unnumbered}
